Single-cell RNA-sequencing reveals cellular dynamics of the developing mammary gland

### 2017-08-24 paper/main.tex
included Johns comments

\normalsize
\textbf{Keywords: Single cell RNAseq, Development, Mammary gland }
\end{titlepage}
\onehalfspacing
\textbf{Abstract:}
\textbf{The mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood.
Characterising the hierarchy of the various mammary epithelial cells and how they are regulated in response to gestation, lactation and involution is important for understanding how breast cancer develops.
Recent studies have used numerous markers to enrich, isolate and characterise the different epithelial cell compartments within the adult mammary gland.
However, in all of these studies only a handful of markers were used to isolate, define and trace cell populations.
Therefore, there is a need for an unbiased and comprehensive description of mammary epithelial cells within the gland at different developmental stages.
To this end we herein used single cell RNA sequencing (scRNAseq) to determine the gene expression profile of individual mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
Our data from 25,010 individual cells identifies 8 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
Interestingly, the effect of gestation and lactation appeared to be more pronounced for some cell types.
For example, our analysis revealed a cluster of luminal progenitor cells in post involution glands, which is distinct from progenitors found in nulliparous glands.
The data also showed that only a few clusters could be fully characterized by a single marker gene.
We argue instead that the epithelial cells -- especially in the luminal compartment -- should rather be conceptualized as being part of a continuous spectrum of differentiation.
This view highlights the plasticity of the tissue and might help to explain some of the conflicting results from lineage tracing studies.}
\newpage

\section{Introduction}
The purpose of the mammary gland is to provide nourishment and passive immunity for the young until they are capable of feeding themselves.
From a developmental biology perspective, the mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood\autocite{Watson2008,Hennighausen2005,Hennighausen1998,Inman2015}.
In the pre-pubertal mouse the mammary gland consists of a rudimentary epithelial ductal structure embedded within a mammary fat pad, which is connected to the nipple\autocite{Mikkola2006,Hens2005}.
At the onset of puberty and in response to hormonal changes, the rudimentary ductal structure will proliferate and migrate to fill the entire mammary fat pad leaving a developed network of ductal structures that later serve as channels for milk transport during lactation.
At the onset of pregnancy a highly proliferative stage is initiated characterised by further ductal side-branching and widespread lobuloalveolar development\autocite{Watson2008}.
Differentiation of the epithelial cells within alveoli prepares the gland for milk production and secretion.
Towards the end of pregnancy the gland is extremely dense and primarily occupied by epithelial cells and very little fat.
This morphology is largely maintained throughout lactation.
However, in response to cessation of suckling the gland undergoes involution, which is characterised by extensive cell death and tissue remodelling \autocite{Inman2015,Watson2006}.
Towards the end of involution the gland reaches a morphology resembling that of glands prior to pregnancy and subsequent pregnancies will trigger the same chain of events.

@@

First, we isolated mammary epithelial cells from four developmental time points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and 11 days post natural involution (PI) (Supplementary Figure 2).
For each time point we sorted mammary epithelial cells based on the EpCAM cell surface marker from two independent mice (\autoref{F1}a).
All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) (Supplementary Figure 3).
Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Figure 4).
First, we dissected the global structure by unsupervised clustering using a shared-nearest-neighbor clustering approach.
This resulted in a coarse clustering into 13 groups (Supplementary Figure 5a).
In a second step we used hierarchical clustering on each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 5b).
After removal of endothelial and immune cell clusters this resulted in a total of 15 clusters of mammary epithelial cells.
Differential gene expression analysis allowed us to further characterise the clusters and infer putative identities (\autoref{F1}d, \autoref{T1} and Supplementary Table 1).
Notably, some clusters mainly represent the same putative cell type at different developmental time points (indicated by the NP, G, L or PI suffix, in the following the suffix might be omitted to refer to a group of clusters). 
Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (Lp, Hsp, Hsd, Avp \& Avd clusters) and 4 a basal profile (Bsl, Myo \& Prc clusters) (\autoref{F1}D, Supplementary Fig 2b and 2c).
The Avd and Avp clusters express high levels of beta-casein, \textit{Csn2}, and are exclusively composed of cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
Hsd and Hsp on the other hand represent the clusters with the highest expression of hormone receptors, suggesting hormone sensing function.
Lp expresses high levels of the luminal progenitor marker \textit{Aldh1a3} suggesting progenitor function, while Hsp and Avp also express progenitor markers but to a lesser extent.
Myo shows characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).

@@

Being confident that the diffusion map recapitulates the luminal differentiation process, we then computationally inferred the two branches and ordered the cells according to their progression through ``pseudotime''\autocite{Haghverdi2016} (see materials and methods for further explanation) (\autoref{F3}a).
This allows us to identify genes whose expression changes during the process of differentiation.
We found 456 genes that showed a pseudotime-dependent expression with the same directionality along both differentiation trajectories (Supplementary Table 2).
These included genes associated with known progenitor characteristics such as \textit{Aldh1a3} and \textit{Tspan8} as well as transcription factors that have not previously been associated with luminal differentiation such as \textit{Creb5}, \textit{Hmga1} and \textit{Fosl1} (\autoref{F3}b).
Interestingly, \textit{Hmga1} is a known chromatin remodeller with reprogramming activity that has been implicated in basal-like breast cancer\autocite{Shah2012,Shah2013}.
In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F3}d, \autoref{F3}e Supplementary Table 3).
We further clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (Supplementary Figure 7).
On the alveolar branch we found two clusters of genes that increase in expression during differentiation, one that is switched on early and another that is activated later in the final stages of differentiation.
The early cluster is enriched for genes involved in fatty-acid oxidation and lipid biosynthetic processes (e.g. \textit{Lipa}, \textit{Dbi} and \textit{Acox1}) .
The second, smaller group of genes that is switched on during the late stages of differentiation was enriched for genes involved in fatty-acid transport and lipid homeostasis (e.g. \textit{Olah}, \textit{Cd36}).
Genes involved in cell division and translation show a transient phase of up-regulation during the alveolar differentiation process (e.g. \textit{Aurka}, \textit{Cenpa}, \textit{Cdk6}), whereas genes regulating cell shape and morphogenesis of a branching epithelium are repressed early on (e.g. \textit{Vdr} and \textit{Areg}).
During the process of differentiation towards hormone-sensing luminal cells, we identified two broad clusters of genes that gradually increase or decrease their expression.
Genes with increasing expression are involved in hormone metabolic processes and the regulation of morphogenesis of an epithelium (e.g. \textit{Vdr}, \textit{Esr1}, \textit{Pgr} and \textit{Msx2}).
Amongst the branch-specific transcription factors we found for example \textit{Runx1, Tox2} and \textit{Bhlhe41} (also known as \textit{Sharp1}) to be transcribed during differentiation towards the hormone-sensing lineage (\autoref{F3}d) and \textit{Elf5, Foxs1} and \textit{Ehf} in the secretory lineage (\autoref{F3}e).
\textit{Runx1} is a known repressor of \textit{Elf5} and its deletion has been shown to be deleterious for ductal morphogenesis\autocite{VanBragt2014}.
The expression of the known progenitor master regulator \textit{Elf5} is maintained and further increased during secretory differentiation suggesting that its transcriptional level is fine-tuned in luminal progenitors (\autoref{F3}e).

@@

Our data illustrates how the cellular composition of the gland changes during the pregnancy cycle.
The luminal compartment shifts from giving rise to mainly hormone sensing cells to producing alveolar, milk-producing cells during pregnancy and lactation. 
The basal compartment on the other hand differentiates to produce oxytocin-sensing, myoepithelial cells that enable duct contraction and milk secretion during lactation.
At the end of lactation the gland then reverts back to a stage that resembles the pre-pregnancy state.
However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}d).
Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like cell types (\autoref{F1}c, Supplementary Figure 8).
To ensure that Lp-PI still represents the progenitor population in the post-parous gland, we identified genes that distinguish Lp-NP from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population Lp-PI.
Indeed, we find genes that are differentially expressed between Lp-NP and the rest of the luminal compartment to show the same trend between Lp-PI and the luminal compartment (\autoref{F4}a).
In a similar manner we can distinguish Lp-PI from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 8b and c).
From this we conclude, that Lp-PI indeed represents the post-parity luminal progenitor population.
Interestingly, genes that were up-regulated in Lp-PI compared to Lp-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}b, Supplementary Fig 8d).
These included genes that play roles in various processes during lactation, e.g. milk-proteins (\textit{Csn2}, \textit{Lalba}), lipases (\textit{Lipa}), proteins involved in milk secretion (\textit{Xdh}, \textit{Cd36}) and transcriptional regulators of lactation (\textit{Cidea}).
Of note, the genes of the casein locus (\textit{Csn2}, \textit{Csn1s1}, \textit{Csn1s2a}, \textit{Csn3}) have previously been reported to be up-regulated in the parous gland, most likely due to changes in chromatin accessibility\autocite{Dos2015,Rijnkels2013}.
However, it has not been shown before that this effect is confined to the progenitor population of the luminal compartment.
Some of the genes that are involved in the immune response are known regulators of the involution process such as \textit{Ctsc}, \textit{Tgfb3} or \textit{Mfge8}.
The up-regulation of these genesets was also present in the Hsp and in some cases even the differentiated Hsd cluster, yet the effect remained strongest in the progenitor cells (\autoref{F4}c).
Finally, we compared the differentiation processes of the luminal compartment of the parous gland to the nulliparous gland by mapping the luminal cells from PI to the trajectory of NP and G cells.
Interestingly, the Hsp and Hsd cluster generally maintain their position of their nulliparous counterpart in the differentiation hierarchy whereas the Lp-PI cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F4}d).
Together, the data suggest that luminal progenitor cells maintain memory of having undergone gestation and involution.
This memory could potentially prime progenitor cells towards the alveolar fate to facilitate alveologenesis in subsequent pregnancies.

@@

We have reported here the use of single cell RNA sequencing to comprehensively map the transcriptomes of thousands of mammary epithelial cells across four developmental time-points.
Our analysis identified 15 clusters of epithelial cells, some of which are only present during specific developmental stage (e.g. Avd and Lp-PI).
This study provides a rich dataset that can be mined online (see link in materials and methods) to identify marker genes and lineage specific promoters that can be used to trace populations of cells \textit{in vivo}.
We note, however, that only some of the clusters can be fully characterized by a single marker gene.
Instead we argue that the epithelial cells - especially in the luminal compartment - should rather be conceptualized as being part of a continuous spectrum of differentiation as visualised in \autoref{F2}.
This view highlights the plasticity of the tissue and might help to explain some of the conflicting results from lineage tracing studies\autocite{Inman2015}.
In this study we could not provide any evidence for contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
We do note, however, that the Prc cluster expresses low levels of both luminal and basal markers, high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
Yet, our data does not support a central role of this cluster in the day-to-day homeostasis of the gland. 
In addition, cells from the Prc cluster also express some but not all markers of pericytes and we thus cannot exclude that these are non-epithelial cells. 
Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (Lp).
Lp gives rise to intermediate states of either hormone-sensing (Hsp) or secretory (Avp) cells that express differentiation markers as well as progenitor markers and appear to represent meta-stable states on the differentiation path from luminal progenitors towards fully differentiated cells (Hsd).
Furthermore, we characterised gene expression patterns along the differentiation hierarchy.
Here we identified genes that are lineage-specific, thus enabling to disentangle the transcriptional events that regulate differentiation of the luminal compartment.
By analysing the mammary gland at various stages of development we were also able to describe the effects of parity at cellular resolution.
We found that the luminal progenitor compartment undergoes lasting changes at the transcriptional level.
This is especially interesting in the light of the protective effect of pregnancies against breast cancer and the role of luminal progenitors as cell of origin.
The majority of the changes were related to pathways involved in immunity and lactation, suggesting that, in particular, the luminal progenitors maintain a memory of gestation and involution.
It is reasonable to assume that Lp-PI overlaps with the previously described parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.

@@

%---------------------------------------------Figure 1--------------------------------------------------------------
\textbf{Figure 1. Single cell RNA sequencing identifies 15 clusters of mammary epithelial cells }
\fakefigure{F1}
\textbf{(a)} Schematic diagram highlighting the experimental setup for isolating and sequencing the RNA of single cells using the 10X chromium system.
\textbf{(b)} t-SNE plot of 25,010 cells visualizes general structure in the data.
Cells are coloured by the four developmental time points Pink=NP, Dark Green=G, Light Green=L, Purple=PI.
\textbf{(c)} Dendrogram of clusters based on the mean expression values of the 15 clusters. The tree was computed based on Spearman's rank correlation with ward linkage.
\textbf{(d)} Same as \textbf{(b)} but colouring cells by clusters.
\textbf{(e)} Heatmap highlighting some key marker genes to distinguish the main classes of clusters.
Colour scale represents log-transformed and normalized UMI counts scaled to a maximum of 1 per row.
The upper panel shows genes that were used to distinguish between luminal and basal cells.
Upper bars represent the cluster assignment and stages for the individual cells.
For visualization purposes only 100 randomly selected cells were shown for large clusters.
%-------------------------------------------------------------------------------------------------------------------

%---------------------------------------------Figure 2--------------------------------------------------------------
\textbf{Figure 2. Computational reconstruction of differentiation processes in the mammary gland}
\fakefigure{F2}
\textbf{(a)} Diffusion map of epithelial cells from the NP and G time points, showing the first three diffusion components.
\textbf{(b)} Differentiation trajectory of the luminal compartment based on the first two diffusion components.
\textbf{(c)} Same plot as in \textbf{(b)}, colored by the normalized and scaled expression values of various genes.
%-------------------------------------------------------------------------------------------------------------------

@@

In total the Cell Ranger software identified 25,806 barcodes that contained enough unique molecules to be considered as cells (4,376 in NP, 6,021 in G, 9,603 in L and 5,806 in PI).
Libraries prepared from all time points showed high quality that was reproducible between the two biological replicates (Supplementary 1a).
We used the following metrics to flag poor quality cells: number of genes detected, total number of unique molecules (UMIs) and percentage of molecules mapped to mitochondrial genes.
Poor quality cells were then identified by setting a threshold on the number of genes and number of UMIs that was defined as three median absolute deviations (MAD) below the median for each sample, while requiring a minimum value of 1,000 total molecules and 500 genes detected.  
This resulted in the following thresholds for total number of genes detected: 1,042 for NP, 836 for G, 500 for L and 759 for PI; and the following for total number of molecules : 2,012 in NP, 1,479 in G, 1,000 in L and 1,379 in G.
In addition, all cells with 5\% or more of UMIs mapping to mitochondrial genes were defined as non-viable or apoptotic and removed from the analysis (Supplementary Fig 1d).
We finally also ensured that none of the reads in our dataset derived from index swapping\autocite{Sinha2017}.
For this we excluded cells from the analysis with barcodes that appeared in more than one sample (non-unique barcodes).
The reasoning being that any index swapped read between samples will also share the same cellular barcode.
Exclusion of these cells hence offers protection against index swapping.
This left us with a total number of 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P).
Gene expression values were then normalised by size factors, that were estimated as previously described\autocite{Lun2016}.
The log-transformed (log2(counts+1)) counts of highly variable genes (HVGs) were used as features for dimensionality reduction and clustering.
HVGs were identified by first fitting a mean-dependent trend to the gene-specific variances to all genes assuming that this trend is dominated by technical variance.
This trend was then used to estimate the technical component of the variance and all genes with a biological component (the residual variance) of at least 0.5 and a Benjamini-Hochberg adjusted P value smaller than or equal to 0.05 were defined as HVG.
The t-SNE embeddings were computed using the ``Rtsne'' package with default settings and perplexity set to 50 (\href{https://github.com/jkrijthe/Rtsne}{\textit{https://github.com/jkrijthe/Rtsne}}).

>>>>

\normalsize
\textbf{Keywords: Single cell RNAseq, Development, Mammary gland }
\end{titlepage}
\onehalfspacing
\textbf{Abstract:}
\textbf{The mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood.
Characterising the hierarchy of the various mammary epithelial cells and how they are regulated in response to gestation, lactation and involution is important for understanding how breast cancer develops.
Recent studies have used numerous markers to enrich, isolate and characterise the different epithelial cell compartments within the adult mammary gland.
However, in all of these studies only a handful of markers were used to define and trace cell populations.
Therefore, there is a need for an unbiased and comprehensive description of mammary epithelial cells within the gland at different developmental stages.
To this end we used single cell RNA sequencing (scRNAseq) to determine the gene expression profile of individual mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
Our data from 25,010 individual cells identifies 8 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
Interestingly, the effect of gestation and lactation appeared to be more pronounced for some cell types.
For example, our analysis revealed a cluster of luminal progenitor cells in post involution glands, which is distinct from progenitors found in nulliparous glands.
The data also showed that few clusters could be fully characterized by a single marker gene.
We argue instead that the epithelial cells -- especially in the luminal compartment -- should rather be conceptualized as being part of a continuous spectrum of differentiation.
This view highlights the plasticity of the tissue and might help to explain some of the conflicting results from lineage tracing studies.}
\newpage

\section{Introduction}
The purpose of the mammary gland is to provide nourishment and passive immunity for the young until they are capable of feeding themselves.
From a developmental biology perspective, the mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood\autocite{Watson2008,Hennighausen2005,Hennighausen1998,Inman2015}.
In the pre-pubertal mouse the mammary gland consists of a rudimentary epithelial ductal structure embedded within a mammary fat pad, which is connected to the nipple\autocite{Mikkola2006,Hens2005}.
At the onset of puberty and in response to hormonal changes, the rudimentary ductal structure will proliferate and migrate to fill the entire mammary fat pad, leaving a developed network of ductal structures that later serve as channels for milk transport during lactation.
At the onset of pregnancy a highly proliferative stage is initiated, characterised by further ductal side-branching and widespread lobuloalveolar development\autocite{Watson2008}.
Differentiation of the epithelial cells within alveoli prepares the gland for milk production and secretion.
Towards the end of pregnancy the gland is extremely dense and primarily occupied by epithelial cells and very little fat.
This morphology is largely maintained throughout lactation.
However, in response to cessation of suckling the gland undergoes involution, which is characterised by extensive cell death and tissue remodelling \autocite{Inman2015,Watson2006}.
Towards the end of involution the gland reaches a morphology resembling that of glands prior to pregnancy and subsequent pregnancies will trigger the same chain of events.

@@

First, we isolated mammary epithelial cells from four developmental time points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and 11 days post natural involution (PI) (Supplementary Figure 2).
For each time point we sorted mammary epithelial cells based on the EpCAM cell surface marker from two independent mice (\autoref{F1}a).
All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) (Supplementary Figure 3).
Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Figure 4).
First, we dissected the global structure by unsupervised clustering using a shared-nearest-neighbor clustering approach.
This resulted in a coarse clustering into 13 groups (Supplementary Figure 5a).
In a second step we applied hierarchical clustering to each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 5b).
After removal of endothelial and immune cell clusters this resulted in a total of 15 clusters of mammary epithelial cells.
Differential gene expression analysis allowed us to further characterise the clusters and infer putative identities (\autoref{F1}d, \autoref{T1} and Supplementary Table 1).
Notably, some clusters mainly represent the same putative cell type at different developmental time points (indicated by the NP, G, L or PI suffix, in the following the suffix might be omitted to refer to a group of clusters). 
Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (Lp, Hsp, Hsd, Avp \& Avd clusters) and 4 a basal profile (Bsl, Myo \& Prc clusters) (\autoref{F1}D, Supplementary Fig 2b and 2c).
The Avd and Avp clusters express high levels of beta-casein, \textit{Csn2}, and are exclusively composed of cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
Hsd and Hsp on the other hand represent the clusters with the highest expression of hormone receptors, suggesting hormone sensing function.
Lp expresses high levels of the luminal progenitor marker \textit{Aldh1a3} suggesting progenitor function, while Hsp and Avp also express progenitor markers but to a lesser extent.
Myo shows characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).

@@

Being confident that the diffusion map recapitulates the luminal differentiation process, we then computationally inferred the two branches and ordered the cells according to their progression through ``pseudotime''\autocite{Haghverdi2016} (see materials and methods for further explanation) (\autoref{F3}a).
This allows us to identify genes whose expression changes during the process of differentiation.
We found 456 genes that showed a pseudotime-dependent expression with the same directionality along both differentiation trajectories (Supplementary Table 2).
These included genes associated with known progenitor characteristics such as \textit{Aldh1a3} and \textit{Tspan8} as well as transcription factors that have not previously been associated with luminal differentiation such as \textit{Creb5}, \textit{Hmga1} and \textit{Fosl1} (\autoref{F3}b).
Interestingly, \textit{Hmga1} is a known chromatin remodeller with reprogramming activity that has been implicated in basal-like breast cancer\autocite{Shah2012,Shah2013}.
In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F3}d, \autoref{F3}e, Supplementary Table 3).
We further clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (\fixme{Supplementary Figure 7}).
On the alveolar branch we found two clusters of genes that increase in expression during differentiation, one that is switched on early and another that is activated later in the final stages of differentiation.
The early cluster is enriched for genes involved in fatty-acid oxidation and lipid biosynthetic processes (e.g. \textit{Lipa}, \textit{Dbi} and \textit{Acox1}) .
The second, smaller group of genes that is switched on during the late stages of differentiation was enriched for genes involved in fatty-acid transport and lipid homeostasis (e.g. \textit{Olah}, \textit{Cd36}).
Genes involved in cell division and translation show a transient phase of up-regulation during the alveolar differentiation process (e.g. \textit{Aurka}, \textit{Cenpa}, \textit{Cdk6}), whereas genes regulating cell shape and morphogenesis of a branching epithelium are repressed early on (e.g. \textit{Vdr} and \textit{Areg}).
During the process of differentiation towards hormone-sensing luminal cells, we identified two broad clusters of genes that gradually increase or decrease their expression.
Genes with increasing expression are involved in hormone metabolic processes and the regulation of morphogenesis of an epithelium (e.g. \textit{Vdr}, \textit{Esr1}, \textit{Pgr} and \textit{Msx2}).
Amongst the branch-specific transcription factors we found for example \textit{Runx1, Tox2} and \textit{Bhlhe41} (also known as \textit{Sharp1}) to be transcribed during differentiation towards the hormone-sensing lineage (\autoref{F3}c) and \textit{Elf5, Foxs1} and \textit{Ehf} in the secretory lineage (\autoref{F3}d).
\textit{Runx1} is a known repressor of \textit{Elf5} and its deletion has been shown to be deleterious for ductal morphogenesis\autocite{VanBragt2014}.
The expression of the known progenitor master regulator \textit{Elf5} is maintained and further increased during secretory differentiation suggesting that its transcriptional level is fine-tuned in luminal progenitors (\autoref{F3}d).

@@

Our data illustrates how the cellular composition of the gland changes during the pregnancy cycle.
The luminal compartment shifts from giving rise to mainly hormone sensing cells to producing alveolar, milk-producing cells during pregnancy and lactation. 
The basal compartment on the other hand differentiates to produce oxytocin-sensing, myoepithelial cells that enable duct contraction and milk secretion during lactation.
At the end of lactation the gland then reverts back to a stage that resembles the pre-pregnancy state.
However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}d).
Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like cell types (\autoref{F1}c, Supplementary Figure 8).
To ensure that Lp-PI still represents the progenitor population in the post-parous gland, we identified genes that distinguish Lp-NP from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population Lp-PI.
Indeed, we find genes that are differentially expressed between Lp-NP and the rest of the luminal compartment to show the same trend between Lp-PI and the luminal compartment (\autoref{F4}a).
In a similar manner we can distinguish Lp-PI from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 8b and c).
From this we conclude, that Lp-PI indeed represents the post-parity luminal progenitor population.
Interestingly, genes that were up-regulated in Lp-PI compared to Lp-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}b, Supplementary Fig 8d).
These included genes that play roles in various processes during lactation, e.g. milk-proteins (\textit{Csn2}, \textit{Lalba}), lipases (\textit{Lipa}), proteins involved in milk secretion (\textit{Xdh}, \textit{Cd36}) and transcriptional regulators of lactation (\textit{Cidea}).
Of note, the genes of the casein locus (\textit{Csn2}, \textit{Csn1s1}, \textit{Csn1s2a}, \textit{Csn3}) have previously been reported to be up-regulated in the parous gland, most likely due to changes in chromatin accessibility\autocite{Dos2015,Rijnkels2013}.
However, it has not been shown before that this effect is confined to the progenitor population of the luminal compartment.
Some of the genes that are involved in the immune response are known regulators of the involution process such as \textit{Ctsc}, \textit{Tgfb3} or \textit{Mfge8}.
The up-regulation of these genesets was also present in the Hsp and in some cases even the differentiated Hsd cluster, yet the effect remained strongest in the progenitor cells (\autoref{F4}c-d).
Finally, we compared the differentiation processes of the luminal compartment of the parous gland to the nulliparous gland by mapping the luminal cells from PI to the trajectory of NP and G cells.
Interestingly, the Hsp and Hsd cluster generally maintain the position of their nulliparous counterpart in the differentiation hierarchy whereas the Lp-PI cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F4}e).
Together, the data suggest that luminal progenitor cells maintain memory of having undergone gestation and involution.
This memory could potentially prime progenitor cells towards the alveolar fate to facilitate alveologenesis in subsequent pregnancies.

@@

We have reported here the use of single cell RNA sequencing to comprehensively map the transcriptomes of thousands of mammary epithelial cells across four developmental time-points.
Our analysis identified 15 clusters of epithelial cells, some of which are only present during specific developmental stage (e.g. Avd and Lp-PI).
This study provides a rich dataset that can be mined online (see link in materials and methods) to identify marker genes and lineage specific promoters that can be used to trace populations of cells \textit{in vivo}.
We note, however, that only some of the clusters can be fully characterized by a single marker gene.
Instead we argue that the epithelial cells - especially in the luminal compartment - should rather be conceptualized as being part of a continuous spectrum of differentiation as visualised in \autoref{F2}.
This view highlights the plasticity of the tissue and might help to explain some of the conflicting results from lineage tracing studies\autocite{Inman2015}.
In this study we could not provide any evidence for contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
We do note, however, that the Prc cluster expresses low levels of both luminal and basal markers, high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
Yet, our data does not support a central role of this cluster in the day-to-day homeostasis of the gland. 
In addition, cells from the Prc cluster also express some but not all markers of pericytes and we thus cannot exclude that these are non-epithelial cells. 
Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (Lp).
Lp gives rise to intermediate states of either hormone-sensing (Hsp) or secretory (Avp) cells that express differentiation markers as well as progenitor markers and appear to represent meta-stable states on the differentiation path from luminal progenitors towards fully differentiated cells (Hsd).
Furthermore, we characterised gene expression patterns along the differentiation hierarchy.
Here we identified genes that are lineage-specific, thus enabling us to disentangle the transcriptional events that regulate differentiation of the luminal compartment.
By analysing the mammary gland at various stages of development we were also able to describe the effects of parity at cellular resolution.
We found that the luminal progenitor compartment undergoes lasting changes at the transcriptional level.
This is especially interesting in the light of the protective effect of pregnancies against breast cancer and the role of luminal progenitors as cell of origin.
The majority of the changes were related to pathways involved in immunity and lactation, suggesting that, in particular, the luminal progenitors maintain a memory of gestation and involution.
It is reasonable to assume that Lp-PI overlaps with the previously described parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.

@@

%---------------------------------------------Figure 1--------------------------------------------------------------
\textbf{Figure 1. Single cell RNA sequencing identifies 15 clusters of mammary epithelial cells }
\fakefigure{F1}
\textbf{(a)} Schematic diagram highlighting the experimental setup for isolating and sequencing the RNA of single cells using the 10X chromium system.
\textbf{(b)} t-SNE plot of 25,010 cells visualizes general structure in the data.
Cells are coloured by the four developmental time points Pink=NP, Dark Green=G, Light Green=L, Purple=PI.
\textbf{(c)} Dendrogram of clusters based on the mean expression values of the 15 clusters. The tree was computed based on Spearman's rank correlation with Ward linkage.
\textbf{(d)} Same as \textbf{(b)} but colouring cells by clusters.
\textbf{(e)} Heatmap highlighting some key marker genes to distinguish the main classes of clusters.
Colour scale represents log-transformed and normalized UMI counts scaled to a maximum of 1 per row.
The upper panel shows genes that were used to distinguish between luminal and basal cells.
Upper bars represent the cluster assignment and stages for the individual cells.
For visualization purposes only 100 randomly selected cells were shown for large clusters.
%-------------------------------------------------------------------------------------------------------------------

%---------------------------------------------Figure 2--------------------------------------------------------------
\textbf{Figure 2. Computational reconstruction of differentiation processes in the mammary gland}
\fakefigure{F2}
\textbf{(a)} Diffusion map of epithelial cells from the NP and G time points, showing the first three diffusion components.
\textbf{(b)} Differentiation trajectory of the luminal compartment based on the first two diffusion components.
\textbf{(c)} Same plot as in \textbf{(b)}, coloured by the normalized and scaled expression values of various genes.
%-------------------------------------------------------------------------------------------------------------------

@@

In total the Cell Ranger software identified 25,806 barcodes that contained enough unique molecules to be considered as cells (4,376 in NP, 6,021 in G, 9,603 in L and 5,806 in PI).
Libraries prepared from all time points showed high quality that was reproducible between the two biological replicates (Supplementary 2a).
We used the following metrics to flag poor quality cells: number of genes detected, total number of unique molecular identifiers (UMIs) and percentage of molecules mapped to mitochondrial genes.
Poor quality cells were then identified by setting a threshold on the number of genes and number of UMIs that was defined as three median absolute deviations (MAD) below the median for each sample, while requiring a minimum value of 1,000 total molecules and 500 genes detected.  
This resulted in the following thresholds for total number of genes detected: 1,042 for NP, 836 for G, 500 for L and 759 for PI; and the following for total number of molecules : 2,012 in NP, 1,479 in G, 1,000 in L and 1,379 in G.
In addition, all cells with 5\% or more of UMIs mapping to mitochondrial genes were defined as non-viable or apoptotic and removed from the analysis (Supplementary Figure 2d).
We finally also ensured that none of the reads in our dataset derived from index swapping\autocite{Sinha2017, Griffiths2017}.
For this we excluded cells with barcodes that appeared in more than one sample (non-unique barcodes).
The reasoning behind this being that any index swapped read between samples will also share the same cellular barcode.
Exclusion of these cells hence offers protection against index swapping.
This left us with a total number of 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P).
Gene expression values were then normalised by size factors, that were estimated as previously described\autocite{Lun2016}.
The log-transformed (log2(counts+1)) counts of highly variable genes (HVGs) were used as features for dimensionality reduction and clustering.
HVGs were identified by first fitting a mean-dependent trend to the gene-specific variances to all genes assuming that this trend is dominated by technical variance.
This trend was then used to estimate the technical component of the variance and all genes with a biological component (the residual variance) of at least 0.5 and a Benjamini-Hochberg adjusted P value smaller than or equal to 0.05 were defined as HVG.
The t-SNE embeddings were computed using the ``Rtsne'' package with default settings and perplexity set to 50 (\href{https://github.com/jkrijthe/Rtsne}{\textit{https://github.com/jkrijthe/Rtsne}}).
